---
title: "Advanced Quantitative Skills in R"
author: "Phillip Hamrick, Ph.D."
date: '2023-01-09'
output: html_document
editor_options: 
  chunk_output_type: console
---

# Data wrangling
Sometimes our data are not laid out right or are hard to read for either you or R. There's lots of things we can do about that. Today, I'm going to show you just a few features that make our life a lot easier in R.


# loading up data from an Excel file and merging data frames
```{r}

# load two data sets, one with experiment data and one with biodata
bio <- read.csv("biodata.csv") # load biodata
wm <- read.csv("working_memory.csv") # load working memory data



# now let's merge these automatically based on matches between participant IDs in each document; we'll use the left_join() function from the dplyr package
library(dplyr)

d <-left_join(bio, wm, by = "Participant") 

```

CAREFUL! Sometimes this step doesn't work because variables aren't labeled correctly as being either a factor or a numerical variable. Let's look at how to change those when we need to.

# converting data types
```{r}

# converting to factor and numeric variable types
# a "factor" variable type is used to represent categorical data - a variable with a limited set of possible values
d$Participant <-as.factor(d$Participant)
# could do d$ParticipantFactor <-as.factor(d$Participant) # this creates a new column on the fly for ya
d$Age <-as.numeric(d$Age)
d$Sex <-as.factor(d$Sex)
d$NativeLanguage <-as.factor(d$NativeLanguage)
d$Working.Memory <-as.numeric(d$Working.Memory)


```

You may want to manipulate other aspects of your data frame, too. Let's look at a couple of common options

```{r}

# dropping unnecessary columns (e.g., what if we wanted our data frame to only have Participant, Age, and WM?)
# we use the subset() function along with the c() function
x <- subset(d, select = c(Participant, Age, Working.Memory)) # keep selected columns
y <- subset(d, select = -c(Sex, NativeLanguage)) # drop selected columns
 


# filtering out unnecessary rows 
#(e.g., what if we didn't want to include Mandarin speakers?)
a <- filter(d, NativeLanguage != "Mandarin")

#(e.g., what if we only wanted Mandarin speakers?)
b <- filter(d, NativeLanguage == "Mandarin")

#can apply multiple filters
mandarinOver25 <- filter(d, 
            NativeLanguage == "Mandarin",
            Age > 25 )

# keep in mind, there are different logical operators in R, and != and == are two of them. you can also use > < and others (e.g., what if you wanted to only look at people older than 21?)
c <- filter(d, Age > 21)


```

A common problem with wrangling data is transposing it from a wide format to a long format and vice versa. Different statistical procedures require different data layouts, pretty much in all statistical software (including SPSS). R is no different. 

# converting from wide to long and vice versa
```{r}

# let's makeup some data in a wide format, which is common for many coders and software (e.g., Qualtrics)
#header=TRUE the first row is headers and not data
olddata_wide <- read.table(header=TRUE, text='
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')

# make sure the subject column is a factor # cast as a factor syntax here factor(tuff to be cast as factor)
olddata_wide$subject <- factor(olddata_wide$subject)

# load the tidyr package
# tidyr, tidyverse, tidymodels <-- good stuffs for peoples commiting to the LONG format!!
install.packages("tidyr")
library(tidyr)

# convert to long format with gather()
# The arguments to gather():
# - data: Data object
# - key: Name of new key column (made from names of data columns)
# - value: Name of new value column
# - ...: Names of source columns that contain values # control:cond2 - everything inbetween control and cond2
# - factor_key: Treat the new key column as a factor (instead of character vector)
data_long <- gather(olddata_wide, condition, measurement, control:cond2, factor_key=TRUE)

#pivot_wider or #pivot_long in dplyr to do a similar thing 

# convert to wide format with spread()
# The arguments to spread():
# - data: Data object
# - key: Name of column containing the new column names
# - value: Name of column containing values
data_wide <- spread(data_long, condition, measurement)

```

```{r}
#subject level data!


#load libraries
library(psych)
library(languageR)

#load data
lexdec <- lexdec

#how many subjects
length(unique(lexdec$Subject))

#descriptive stats - descriptive stats for each person ! - Clunky AF - can't plop this into dataframe
dataframe<-describeBy(lexdec$RT, group= lexdec$Subject)

#aggregate into a dataframe subject level means for reaction time
#can use a different function (FUN) . Can use median or what ever else rather then mean
subj = aggregate(RT ~ Subject, data =lexdec, FUN=mean)

#add in more information from our data set
subj2 = aggregate(RT ~ Subject + Sex + NativeLanguage, data =lexdec, FUN=mean)

#add in conditions? animal or plant
subj3 = aggregate(RT ~ Subject + Sex + NativeLanguage + Class, data =lexdec, FUN=mean)

#when yo aggregate a dataset , plz rename it lol

```

```{r}
#create a graph from data frame d that is only female participants working memory data and look also at their native language

#assign colors to native languanges
#use facet wrap to have each naive language in a graph

dLadiesOnly = filter(d, Sex == "0",
                     NativeLanguage == "Arabic" | NativeLanguage =="Mandarin")
library(ggplot2)
ggplot(dLadiesOnly, aes (y=Working.Memory, x=NativeLanguage, fill= NativeLanguage)) +geom_boxplot(alpha =.60) + geom_jitter(alpha=0.5)

#you can set the colors for ggplot , you can also have additional packages that let you color ggplot somemore


```

